---
title: "README"
output: html_document
date: "2023-07-24"
---

## Background

This manuscript provides context and details about the smdi R package that was developed as a result of the findings of a comprehensive Sentinel Innovation Center project on `principled approaches on partially observed confounder data` in electronic health records.

More information about the package can be found under:
<https://janickweberpals.gitlab-pages.partners.org/smdi>

## Structure

-   .Rprofile - defines paths and activates `renv`, options for Posit R package manager
-   tables - main and supplementary tables (R objects and .docx format)
-   figures - main and supplementary figures (R objects and .docx format)
-   renv/renv.lock - `renv` directories to manage R dependencies and versions used in this simulation

## Dependencies

R package dependencies are managed through the `renv` package. All packages and their versions can be viewed in the lockfile `renv.lock`. All required packages and the appropriate versions can be installed by running the following command: 

```{r, eval=FALSE}
renv::restore()
```

## Directory overview

```{r}
fs::dir_tree(recurse = 1)
```

```{r, include=FALSE, eval=FALSE}
rmarkdown::render(input = "README.Rmd", output_format = "md_document")
```